---
title: "Market Segmentation"
author: "Sejal Deepak Kankriya"
date: '2024-02-20'
output:
  pdf_document: default
  html_document:
    df_print: paged
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(cluster)
library(dplyr)
library(magrittr)
library(ggplot2)
library(plotly)
library(data.table)
library(caret)
library(ggbiplot)
library(tidyr)
library(cowplot)
library(ggdendro)
library(factoextra)
```

# Market Segmentation

An advertisement division of large club store needs to perform customer
analysis the store customers in order to create a segmentation for more
targeted marketing campaign

The task is to identify similar customers and characterize them (at
least some of them). In other word perform clustering and identify
customers segmentation.

This data-set is derived from
<https://www.kaggle.com/imakash3011/customer-personality-analysis>

```         
Colomns description:
People
  ID: Customer's unique identifier
  Year_Birth: Customer's birth year
  Education: Customer's education level
  Marital_Status: Customer's marital status
  Income: Customer's yearly household income
  Kidhome: Number of children in customer's household
  Teenhome: Number of teenagers in customer's household
  Dt_Customer: Date of customer's enrollment with the company
  Recency: Number of days since customer's last purchase
  Complain: 1 if the customer complained in the last 2 years, 0 otherwise

Products

  MntWines: Amount spent on wine in last 2 years
  MntFruits: Amount spent on fruits in last 2 years
  MntMeatProducts: Amount spent on meat in last 2 years
  MntFishProducts: Amount spent on fish in last 2 years
  MntSweetProducts: Amount spent on sweets in last 2 years
  MntGoldProds: Amount spent on gold in last 2 years

Place
  NumWebPurchases: Number of purchases made through the company’s website
  NumStorePurchases: Number of purchases made directly in stores
```

Assume that data was current on 2014-07-01

Read Dataset and Data Conversion to Proper Data Format

Read "m_marketing_campaign.csv" using `data.table::fread` command,
examine the data.

```{r}
# fread m_marketing_campaign.csv and save it as df
# Read the data
df <- fread("m_marketing_campaign.csv")
# Examine the data
head(df)
```

```{r}
# Convert Year_Birth to Age (assume that current date is 2014-07-01)
df[, Age := 2014 - Year_Birth]
df$Age <- as.integer(df$Age)

# Dt_Customer is a date (it is still character), convert it to membership days (i.e. number of days person is a member, name it MembershipDays)
# Convert Dt_Customer to MembershipDays

df[, MembershipDays := as.Date("2014-07-01", format = "%Y-%m-%d") - as.Date(Dt_Customer, format = "%d-%m-%Y")]
df$MembershipDays <- as.integer(df$MembershipDays)
```

```{r}
# Summarize Education column (use table function)
table(df$Education)

# Lets treat Education column as ordinal categories and use simple levels for distance calculations
# Assuming following order of degrees:
#    HighSchool, Associate, Bachelor, Master, PhD
# factorize Education column (hint: use factor function with above levels)
df$Education <- factor(df$Education, levels = c("HighSchool", "Associate", "Bachelor", "Master", "PhD"))
```

```{r}
# Summarize Marital_Status column (use table function) 
table(df$Marital_Status)

# Lets convert single Marital_Status categories for 5 separate binary categories
# Divorced, Married, Single, Together and Widow, the value will be 1 if customer 
# is in that category and 0 if customer is not
# use dummyVars from caret package, model.matrix or simple comparison (there are only 5 groups)

df$Marital_Status <- factor(df$Marital_Status)
dummy <- dummyVars(~ Marital_Status, data = df)
dummies_conv <- data.frame(predict(dummy, newdata = df))

# dummies
colnames(dummies_conv) <- c("Divorced", "Married", "Single", "Together", "Widow")
df <- cbind(df, dummies_conv)
```

```{r}
# lets remove columns which we will no longer use:
# remove ID, Year_Birth, Dt_Customer, Marital_Status
# and save it as df_sel 
# df_sel <- df %>% select(-ID, -Year_Birth, -Dt_Customer, -Marital_Status)

df_sel <- subset(df, select = -c(ID, Year_Birth, Dt_Customer, Marital_Status))

# Convert Education to integers 
# use as.integer function, if you use factor function earlier 
# properly then HighSchool will be 1, Associate will be 2 and so on)

df_sel$Education <- as.integer(df_sel$Education)
# head(df_sel)
```

```{r}
# lets scale
# run scale function on df_sel and save it as df_scale
# that will be our scaled values which we will use for analysis
df_scale <- scale(df_sel)
# head(df_scale)
```

## PCA

Run PCA

```{r}
# Run PCA on df_scale, make biplot and scree plot/percentage variance explained plot
# save as pc_out, we will use pc_out$x[,1] and pc_out$x[,2] later for plotting

pc_out <- prcomp(df_scale, scale = TRUE)
# pc_out

plot(pc_out, type="line", main="Scree Plot")
```

```{r}
# df_scale_hc <- hclust(dist(df_scale), method='complete')
# df_ct <- cutree(df_scale_hc, k=5)

par(mfrow = c(1, 2))
plot(x=pc_out$x[,1],y=pc_out$x[,2], col = df$NumStorePurchases, pch = 19,
    xlab = "P1", ylab = "P2", main = "Store Purchase Analysis")
plot(x=pc_out$x[,1],y=pc_out$x[,2], col = df$NumWebPurchases, pch = 19,
    xlab = "P1", ylab = "P2", main = "Online Purchase Analysis")
```

```{r}
# Biplot
ggbiplot(pc_out, scale=0, groups = as.factor(df$Marital_Status))
```

```{r}

# percentage variance explained plot

pve <- 100 * pc_out$sdev^2 / sum(pc_out$sdev^2)
par(mfrow = c(1, 2))
plot(pve,  type = "o", ylab = "PVE",
    xlab = "Principal Component", col = "blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE",
    xlab = "Principal Component", col = "brown3")
```

The variables related to customer purchases (NumStorePurchases and
NumWebPurchases) do not show any clear clusters or patterns. However, we
can see some patterns in the variable Recency, which is a measure of how
many days have passed since the customer's last purchase. According to
the plotly function, there are three distinct groups of customers based
on Recency, with different colors representing each group.

Furthermore, the scree plot shows that the first principal component
explains more than 30% of the total variance, while the second principal
component explains only about 20%. The first two principal components
account for more than half of the total variance. This shows that these
two components may be capturing some underlying elements or variables
that are driving customer behavior. However, further research is
required to discover these elements and comprehend how they relate to
client purchases and other variables in the dataset.

## Cluster with K-Means

use K-Means method for clustering

### Selecting Number of Clusters

Select optimal number of clusters using elbow method.

```{r}
set.seed(2)
km_out_list <- lapply(1:10, function(k) list(
  k=k,
  km_out=kmeans(df_scale, k, nstart = 20)))

km_results <- data.frame(
  k=sapply(km_out_list, function(k) k$k),
  totss=sapply(km_out_list, function(k) k$km_out$totss),
  tot_withinss=sapply(km_out_list, function(k) k$km_out$tot.withinss)
  )
km_results
plot_ly(km_results,x=~k,y=~tot_withinss) %>% add_markers() %>% add_paths()
```

```{r}

# Elbow method
set.seed(4)
fviz_nbclust(df_scale, kmeans, method = "wss",k.max=10, nstart=20, iter.max=20) +
  geom_vline(xintercept = 2, linetype = 2)+
  labs(subtitle = "Elbow method")
```

Select optimal number of clusters using Gap Statistic.

```{r}
set.seed(3)
gap_kmeans <- clusGap(df_scale, kmeans, nstart = 20, K.max = 10, B = 100, iter.max=30)
plot(gap_kmeans, main = "Gap Statistic: kmeans")
```

```{r}

# Gap statistic
fviz_nbclust(df_scale, kmeans, method = "gap_stat", nboot = 20,k.max=20, nstart=20, iter.max=40) +
  labs(subtitle = "Gap statistic method")
```

Select optimal number of clusters using Silhouette method.

```{r fig.width=8,fig.height=8}

set.seed(5)

#Silhouette
par(mar = c(5, 2, 4, 2), mfrow=c(2,2))
for(k in c(2,3,4,9)) {
  kmeans_ct <- kmeans(df_scale, k, nstart=20)
  si <- silhouette(kmeans_ct$cluster, dist = dist(df_scale))
  plot(si, main="")
}
par(mar = c(1, 1, 1, 1), mfrow=c(1,1))
```

```{r fig.width=9,fig.height=9}

set.seed(1)

#Silhouette
results <- lapply(2:20, function(k) {
  kmeans_cluster <- kmeans(df_scale, k, nstart=20)
  si <- silhouette(kmeans_cluster$cluster, dist = dist(df_scale))
  data.frame(k=k,sil_width=mean(si[,'sil_width']),sil_width_min=min(si[,'sil_width']))
})
si_df <- bind_rows(results)

plot_ly(si_df, x=~k,y=~sil_width) %>%
  add_markers() %>% add_lines() %>%
  add_markers(y=~sil_width_min) %>% add_lines(y=~sil_width_min)

ggplot(si_df, aes(x=k,y=sil_width,color="Width Avg"))+geom_point()+geom_line()+
  geom_point(aes(y=sil_width_min,color="Width Min"))+geom_line(aes(y=sil_width_min,color="Width Min"))
```

```{r}
fviz_nbclust(df_scale, kmeans, method = "silhouette", nboot = 20,k.max=20, nstart=20, iter.max=40)+
  labs(subtitle = "Silhouette method")
```

Deciding which k to choose based on elbow, gap statistics and silhuettes
as well as clustering task (market segmentation for advertisement
purposes, that is two groups don't provide sufficient benefit over a
single groups).

It has been concluded that the ideal number of k is two, according to
the elbow method, the gap method, and the silhouette method. However, it
is not enough to segment customers into two segments. A clustering
should be conducted with 3 as a value of k since, based on the elbow
plot, gap statistics, and Silhouette methods, this is the second-most
optimal number of k.

It is stated that two groups do not provide sufficient benefit over a
single group for the market segmentation task, implying that at least
three groups are required to obtain meaningful segmentation. As a
result, based on the elbow, gap statistics, and silhouette results, as
well as the clustering task requirements, it is recommended to select
three clusters for the market segmentation analysis.

## Clusters Visulalization

Make k-Means clusters with selected k_kmeans (store result as km_out).
Plot your k_kmeans clusters on biplot (just PC1 vs PC2) by coloring
points by their cluster id.

```{r}

# Computing kmeans with new selected k value

set.seed(6)
km_out <- kmeans(df_scale, 3, nstart = 25)
# km_out
```

```{r}
fviz_cluster(km_out, data = df_scale, ellipse.type = "euclid", star.plot = TRUE,
             repel = TRUE, ggtheme = theme_classic())
```

The k-means clustering algorithm identified three distinct groups based
on the plot. The red cluster is located in the plot's upper right corner
and is distinguished by high values of the variables represented by the
x and y axes. This category may include customers who are high spenders
and make frequent purchases.

The green cluster is located in the plot's bottom left corner and is
distinguished by low values of the variables represented by the x and y
axes. Customers in this category may be low spenders who make infrequent
purchases.

The blue cluster is located in the top left corner of the plot and is
distinguished by high x-axis values and low y-axis values. Customers in
this category may be high spenders who make infrequent purchases.

## Characterizing Cluster

Perform descriptive statistics analysis on obtained cluster. Based on
that does one or more group have a distinct characteristics.

```{r}
# df$cluster <- as.factor(km_out$cluster)
df_sel <- df_sel %>%
  mutate(cluster = km_out$cluster)

stat_analysis <- df_sel %>%
  group_by(cluster) %>%
  summarise_all(list(mean = mean, sd = sd))

print(stat_analysis)
```

## Cluster with Hierarchical Clustering

Perform clustering with Hierarchical method. Try complete, single and
average linkage. Plot dendagram, based on it choose linkage and number
of clusters.

```{r}
set.seed(6)
distance_matrix <- dist(df_scale, method = "euclidean")
hc_complete <- hclust(distance_matrix, method = "complete")
hc_avg <- hclust(distance_matrix, method = "average")
hc_single <- hclust(distance_matrix, method = "single")

par(mfrow = c(1, 3))
plot(hc_complete, main = "Complete Linkage",
    xlab = "", sub = "", cex = .3, hang = -1)
plot(hc_avg, main = "Average Linkage",
    xlab = "", sub = "", cex = .3, hang = -1)
plot(hc_single, main = "Single Linkage",
    xlab = "", sub = "", cex = .3, hang = -1)
```

```{r}
# ggplot
ggdendrogram(hc_complete, segements=TRUE, labels=TRUE, leaf_labels = TRUE, rotate=FALSE, theme_dendro = TRUE) +
 labs(title='Complete Linkage')

ggdendrogram(hc_avg, segements=TRUE, labels=TRUE, leaf_labels = TRUE, rotate=FALSE, theme_dendro = TRUE) +
 labs(title='Average Linkage')

ggdendrogram(hc_single, segements=TRUE, labels=TRUE, leaf_labels = TRUE, rotate=FALSE, theme_dendro = TRUE) +
 labs(title='Single Linkage')
```

In my opinion, scaling is essential in order to avoid bias in
conclusions. In the dendrograms, the maximum number of distinct clusters
is generated by the method of complete linkage, whereas the minimum
number of distinct clusters is generated by the method of single
linkage. Clusters with an average linkage are located in the middle.

If there are elbows in the data, this indicates a natural break in the
data, which can be used to determine the number of clusters. A slight
elbow can be observed around three clusters using the full and average
linkage methods. In contrast, the single linkage method does not
indicate a clear correlation between the two.
